### 2018-11-02 pkg/model4you/inst/JORS/man_model4you.Rnw
consistent font of R code in text and chunks
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2373 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b971c2b79875d0ea5072932a285afc40ae63a35d

Treatment effect estimation is often done using simple models with the binary
treatment indicator as only covariate. In the example of a clinical trial the
treatment indicator would be $1$ if the patient receives the new treatment and
$0$ if the patient receives standard of care.  In R such a simple model can be
estimated as follows:
<<eval=FALSE>>=
base_model <- model(response ~ treatment, data)
@
with \Sexpr{'response'} being the response measured, \Sexpr{'treatment'} being
the treatment indicator and \Sexpr{'data'} being the data set containing these
variables.  The function \Sexpr{'model()'} can be replaced for example by
\Sexpr{'lm()'} to estimate a linear model, \Sexpr{'glm()'} to estimate a
generalised linear model or \Sexpr{'survreg()'} to estimate a parametric
survival model.  These models estimate intercept and treatment effect for all
individuals in the \Sexpr{'data'} and allow for predicting the response of
other individuals given they do or do not receive the treatment of interest.

@@

\paragraph{1. Model-based trees} identify subgroups where within the subgroups
the model parameters are similar and between groups the model parameters are
different. This is achieved by finding instabilities in the model parameters
with respect to a variable (characteristic) and recursively partitioning the
data into groups. If, for example the algorithm finds that men and women have
differing treatment effects, the data are partitioned into two subgroups.
Details on model-based trees in general can be found in
\citep{zeileis_model-based_2008} and for the special use case for stratified
treatment effect estimation in \citep{seibold_model-based_2016}.  Just a single
line of code lets the user compute a model-based tree in R:
<<eval=FALSE>>=
strat_models <- pmtree(base_model)
@
Note that \Sexpr{'pmtree()'} uses the data given in the call of the base model.
It automatically uses variables not used in the model formula (in the example
above \Sexpr{'response ~ treatment'}) as potential subgroup defining variables.
This can be edited using the \Sexpr{'zformula'} argument.

\paragraph{2. Personalised models} use model-based random forests to estimate
similarity of individuals in terms of model parameters. For each individual a
personalised model can be estimated based on a weighted set of the original
data, where the similarity measure corresponds to the weight. Details on the
personalised models can be found in \citep{seibold_individual_2017}.  Computing
personalised models for all observations in the training data is simple:
<<eval=FALSE>>=
pm_forest <- pmforest(base_model)
pers_models <- pmodel(pm_forest)
@
Again here the potential effect-modifying variables are taken by default as all
variables not given in the model formula and can be defined using the
\Sexpr{'zformula'} argument in \Sexpr{'pmforest()'}.

@@

A tree based on this model can be computed and visualised in only two lines of
code:
<<math_tree, fig.width = 13, fig.cap = "Personalised model tree for the Mathematics Exam datam.">>=
tr_math <- pmtree(bmod_math, control = ctree_control(maxdepth = 2))
plot(tr_math, terminal_panel = node_pmterminal(tr_math,
                                               plotfun = lm_plot))
@
\new{
We restrict the depth of the tree to two (\Sexpr{'maxdepth = 2'}) for
illustration purposes. If this setting is not used a more complex tree would be
estimated. Also the cutpoints and effect estimates are rounded in the plot.  To
view the tree in the R console, the command \Sexpr{'print()'} function can be
used.
}
<<math_printtree>>=
print(tr_math)
@
The tree (see Figure~\ref{fig:math_tree}) divides students based on the
percentage of successful online \Sexpr{'tests'}. These online tests were
conducted biweekly throughout the semester.  The largest difference between the
two exam groups is in the students who did very well in the online tests (more
than \Sexpr{round(tr_math[[5]]$node$split$breaks, 1)} percent correct). The
tree thus gives us much more information on the fairness of the exam than the
simple linear model, which is that it does not seem to be fair for students who
did very well throughout the semester (at this point we should state that the
students self selected into the two exam groups which might also be the reason
for differences in exam performance).
\new{Note that confidence intervals shown in the tree are not adjusted and
should hence be interpreted as a measure for variablity and not be used for
inference.}

@@

Dependence plots with the group effect (treatment effect) on the y-axis and
the student characteristics on the x-axis are a good way of visualising the
personalised models and for getting knowledge about the interactions between
student characteristics and the treatment.
\new{
Note that the plot is related to but not the same as the classical partial
dependence plot \citep{friedman2001greedy}, as each point in the figure is one
patient in the given data set.
}
Since the percentage of successful online tests is measured on a grid, a
\new{beeswarm} plot possibly shows the relationships even better than the
scatter plot (both shown in Figure~\ref{fig:math_dp1},
\new{
Figure~\ref{fig:math_dp11} includes a loess curve). We see a nonlinear relation
between the percentage of successful online \Sexpr{'tests'} and the exam group
effect. Students with great performance during the semenster are estimated to
have the strongest effect.
}
<<math_dp1, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plot for percentage of tests successfully solved.", fig.subcap=c("Scatter plot with loess curve.", "Beeswarm plot.")>>=
dpdat_math <- cbind(pmods_math, MathExam)

@@

\new{
For categorical variables such as the number of previous attempts to pass the
exam (\Sexpr{'attempt'}) and \Sexpr{'gender'} we can use box plots,
\new{beeswarm} plots or a combination thereof can be used (as in
Figure~\ref{fig:math_dp21}). Figure~\ref{fig:math_dp22} shows the number of
previous attempts on the x-axis and the color indicates the gender of each
student.  This way we can see that for students writing the exam for the first
time, gender seems to play a bigger role in terms of estimated individual exam
group effect than for others.
}
<<math_dp2, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plots for the number of previous attempts and gender.", fig.subcap=c('Gender.', 'Number of previous attempts and gender.')>>=
ggplot(dpdat_math, aes(x = gender, y = group2, color = gender)) +
  geom_boxplot() +
  geom_quasirandom(alpha = 0.5) +
  theme(legend.position = "none") +
  ylab("estimated individual\nexam group effect")

@@

\section*{Implementation and architecture}
The R package \emph{model4you} is focused on ease of use and
interpretability.  Users can take a simple model that they know and
understand as basis.
\new{
Models currently available for use are linear models (function
\Sexpr{'lm()'} in R), generalised linear models (\Sexpr{'glm()'}) and
survival models (\Sexpr{'survreg()'} and \Sexpr{'coxph()'} from the R
package \emph{survival}).  Models are restricted to those with a
single binary covariate (e.g.~ two treatment options).  The user can
simply plug the model object into \Sexpr{'pmtree()'} or
\Sexpr{'pmforest()'} depending on whether they want subgroup wise or
personalised models.
}

The basis for these functionalities is provided by the
\emph{partykit} package which is a widely used R package for trees
and forests \citep{hothorn_partykit_2015, hothorn_partykit_2017}.
The \emph{model4you} package provides wrappers for the well
implemented and tested functions \Sexpr{'partykit::ctree()'} and
\Sexpr{'partykit::cforest()'} and extends the functionalities to
allow for the computation of personalised models and to improve
usability and interpretability.

@@

The software is intentionally written to make usage as simple as possible. The
most prominent use case are clinical trials where the assumption of an average
treatment effect for all patients is too strict and the efficacy of the
treatment depends on patient characteristics (e.g.~ gender, biomarkers, etc.).
For subgroup analyses (stratified treatment effects) model-based trees
(\Sexpr{'pmtree()'}) can be used; For personalised treatment effects
model-based forests (\Sexpr{'pmforest()'}) provide a way of estimating
similarity between patients and using this similarity measure to estimate
personalised models (\Sexpr{'pmodel()'}).  The target audience are people who
deal with heterogeneous treatment effects, such as medical researchers,
pharmaceutical companies or analysts in marketing (A-B testing).  In general
the software is useful to researchers dealing with scenarios where two
exposures are compared and responses of subjects possibly depend on other
variables.

>>>>

Treatment effect estimation is often done using simple models with the binary
treatment indicator as only covariate. In the example of a clinical trial the
treatment indicator would be $1$ if the patient receives the new treatment and
$0$ if the patient receives standard of care.  In R such a simple model can be
estimated as follows:
<<eval=FALSE>>=
base_model <- model(response ~ treatment, data)
@
with \texttt{\Sexpr{'response'}} being the response measured, \texttt{\Sexpr{'treatment'}} being
the treatment indicator and \texttt{\Sexpr{'data'}} being the data set containing these
variables.  The function \texttt{\Sexpr{'model()'}} can be replaced for example by
\texttt{\Sexpr{'lm()'}} to estimate a linear model, \texttt{\Sexpr{'glm()'}} to estimate a
generalised linear model or \texttt{\Sexpr{'survreg()'}} to estimate a parametric
survival model.  These models estimate intercept and treatment effect for all
individuals in the \texttt{\Sexpr{'data'}} and allow for predicting the response of
other individuals given they do or do not receive the treatment of interest.

@@

\paragraph{1. Model-based trees} identify subgroups where within the subgroups
the model parameters are similar and between groups the model parameters are
different. This is achieved by finding instabilities in the model parameters
with respect to a variable (characteristic) and recursively partitioning the
data into groups. If, for example the algorithm finds that men and women have
differing treatment effects, the data are partitioned into two subgroups.
Details on model-based trees in general can be found in
\citep{zeileis_model-based_2008} and for the special use case for stratified
treatment effect estimation in \citep{seibold_model-based_2016}.  Just a single
line of code lets the user compute a model-based tree in R:
<<eval=FALSE>>=
strat_models <- pmtree(base_model)
@
Note that \texttt{\Sexpr{'pmtree()'}} uses the data given in the call of the base model.
It automatically uses variables not used in the model formula (in the example
above \texttt{\Sexpr{'response ~ treatment'}}) as potential subgroup defining variables.
This can be edited using the \texttt{\Sexpr{'zformula'}} argument.

\paragraph{2. Personalised models} use model-based random forests to estimate
similarity of individuals in terms of model parameters. For each individual a
personalised model can be estimated based on a weighted set of the original
data, where the similarity measure corresponds to the weight. Details on the
personalised models can be found in \citep{seibold_individual_2017}.  Computing
personalised models for all observations in the training data is simple:
<<eval=FALSE>>=
pm_forest <- pmforest(base_model)
pers_models <- pmodel(pm_forest)
@
Again here the potential effect-modifying variables are taken by default as all
variables not given in the model formula and can be defined using the
\texttt{\Sexpr{'zformula'}} argument in \texttt{\Sexpr{'pmforest()'}}.

@@

A tree based on this model can be computed and visualised in only two lines of
code:
<<math_tree, fig.width = 13, fig.cap = "Personalised model tree for the Mathematics Exam datam.">>=
tr_math <- pmtree(bmod_math, control = ctree_control(maxdepth = 2))
plot(tr_math, terminal_panel = node_pmterminal(tr_math,
                                               plotfun = lm_plot))
@
\new{
We restrict the depth of the tree to two (\texttt{\Sexpr{'maxdepth = 2'}}) for
illustration purposes. If this setting is not used a more complex tree would be
estimated. Also the cutpoints and effect estimates are rounded in the plot.  To
view the tree in the R console, the command \texttt{\Sexpr{'print()'}} function can be
used.
}
<<math_printtree>>=
print(tr_math)
@
The tree (see Figure~\ref{fig:math_tree}) divides students based on the
percentage of successful online \texttt{\Sexpr{'tests'}}. These online tests were
conducted biweekly throughout the semester.  The largest difference between the
two exam groups is in the students who did very well in the online tests (more
than \texttt{\Sexpr{round(tr_math[[5]]$node$split$breaks, 1)}} percent correct). The
tree thus gives us much more information on the fairness of the exam than the
simple linear model, which is that it does not seem to be fair for students who
did very well throughout the semester (at this point we should state that the
students self selected into the two exam groups which might also be the reason
for differences in exam performance).
\new{Note that confidence intervals shown in the tree are not adjusted and
should hence be interpreted as a measure for variablity and not be used for
inference.}

@@

Dependence plots with the group effect (treatment effect) on the y-axis and
the student characteristics on the x-axis are a good way of visualising the
personalised models and for getting knowledge about the interactions between
student characteristics and the treatment.
\new{
Note that the plot is related to but not the same as the classical partial
dependence plot \citep{friedman2001greedy}, as each point in the figure is one
patient in the given data set.
}
Since the percentage of successful online tests is measured on a grid, a
\new{beeswarm} plot possibly shows the relationships even better than the
scatter plot (both shown in Figure~\ref{fig:math_dp1},
\new{
Figure~\ref{fig:math_dp11} includes a loess curve). We see a nonlinear relation
between the percentage of successful online \texttt{\Sexpr{'tests'}} and the exam group
effect. Students with great performance during the semenster are estimated to
have the strongest effect.
}
<<math_dp1, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plot for percentage of tests successfully solved.", fig.subcap=c("Scatter plot with loess curve.", "Beeswarm plot.")>>=
dpdat_math <- cbind(pmods_math, MathExam)

@@

\new{
For categorical variables such as the number of previous attempts to pass the
exam (\texttt{\Sexpr{'attempt'}}) and \texttt{\Sexpr{'gender'}} we can use box plots,
\new{beeswarm} plots or a combination thereof can be used (as in
Figure~\ref{fig:math_dp21}). Figure~\ref{fig:math_dp22} shows the number of
previous attempts on the x-axis and the color indicates the gender of each
student.  This way we can see that for students writing the exam for the first
time, gender seems to play a bigger role in terms of estimated individual exam
group effect than for others.
}
<<math_dp2, fig.width=4, fig.height=3, out.width='.48\\linewidth', fig.cap="Dependence plots for the number of previous attempts and gender.", fig.subcap=c('Gender.', 'Number of previous attempts and gender.')>>=
ggplot(dpdat_math, aes(x = gender, y = group2, color = gender)) +
  geom_boxplot() +
  geom_quasirandom(alpha = 0.5) +
  theme(legend.position = "none") +
  ylab("estimated individual\nexam group effect")

@@

\section*{Implementation and architecture}
The R package \emph{model4you} is focused on ease of use and
interpretability.  Users can take a simple model that they know and
understand as basis.
\new{
Models currently available for use are linear models (function
\texttt{\Sexpr{'lm()'}} in R), generalised linear models (\texttt{\Sexpr{'glm()'}}) and
survival models (\texttt{\Sexpr{'survreg()'}} and \texttt{\Sexpr{'coxph()'}} from the R
package \emph{survival}).  Models are restricted to those with a
single binary covariate (e.g.~ two treatment options).  The user can
simply plug the model object into \texttt{\Sexpr{'pmtree()'}} or
\texttt{\Sexpr{'pmforest()'}} depending on whether they want subgroup wise or
personalised models.
}

The basis for these functionalities is provided by the
\emph{partykit} package which is a widely used R package for trees
and forests \citep{hothorn_partykit_2015, hothorn_partykit_2017}.
The \emph{model4you} package provides wrappers for the well
implemented and tested functions \texttt{\Sexpr{'partykit::ctree()'}} and
\texttt{\Sexpr{'partykit::cforest()'}} and extends the functionalities to
allow for the computation of personalised models and to improve
usability and interpretability.

@@

The software is intentionally written to make usage as simple as possible. The
most prominent use case are clinical trials where the assumption of an average
treatment effect for all patients is too strict and the efficacy of the
treatment depends on patient characteristics (e.g.~ gender, biomarkers, etc.).
For subgroup analyses (stratified treatment effects) model-based trees
(\texttt{\Sexpr{'pmtree()'}}) can be used; For personalised treatment effects
model-based forests (\texttt{\Sexpr{'pmforest()'}}) provide a way of estimating
similarity between patients and using this similarity measure to estimate
personalised models (\texttt{\Sexpr{'pmodel()'}}).  The target audience are people who
deal with heterogeneous treatment effects, such as medical researchers,
pharmaceutical companies or analysts in marketing (A-B testing).  In general
the software is useful to researchers dealing with scenarios where two
exposures are compared and responses of subjects possibly depend on other
variables.
